### 2018-11-15 Manuscript/Manuscript.tex

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering. A systematic analysis of LIP area through the Phanerozoic reveals that there is no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

This chapter seeks to address two interconnected questions: 1) how unique are the peaks in low-latitude LIP area that have been proposed to be associated with climatic cooling? and 2) how strong is the overall relationship between tropical LIP area and glaciation?

@@

Outlines of LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were taken from the compilation of \citet{Ernst2017a}, and emplacement ages were taken from the literature (Fig. \ref{fig:LIP_map}; Table \ref{tab:LIPs}). The compilation of \citet{Ernst2017a} seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the construction of the original extent polygons by \citet{Ernst2017a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude the 6 of 19 LIPs which have been partially or completely buried in our estimation of a representative LIP area half-life. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the root mean square error of the fit is low (0.14), there is variability in the half-lives for LIPs if they are fit individually with an exponential decay function, such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36~Myr' uses the best-fit half-life of 36~Myr, while `t$_{1/2}$ = 120~Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly impact the calculated LIP areas throughout the Phanerozoic, which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial by sediments of LIP volcanics that are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts as well as successful rifting in the region of the Red Sea that has resulted in burial (Fig \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios.

@@

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and Paleozoic Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. The total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree~S to 15\textdegree~N) were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_area}B and C). We use 15\textdegree~S to 15\textdegree~N as the definition of the tropical rain belt as those latitudes approximately correspond with precipitation greater than 1.0~m/yr in modern climatalogical data \citep{Kalnay1996a}. The consistent paleolatitudes of evaporite deposits in the subtropics demonstrate the stability of the large-scale atmospheric circulation that gives rise to intense precipitation in the tropical rain belt \citep{Evans2006a}.

@@

To assess the statistical significance of the correlation (or lack thereof) implied by the Pearson correlation coefficients, we simulated the four glacial episodes (Fig. \ref{fig:LIP_area}E) occurring at random times through the past 520~Myr, and recomputed the correlation coefficient and \% overlap between the LIP area in the tropics and the randomly timed glacial intervals for each of these 100,000 simulations (Fig. \ref{fig:LIP_correlation}). This approach accounts for the reality that spurious correlation can arise between auto-correlated data sets such as these, where each value is not independent, but is instead dependent on the previous state of the system. For the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario, 28\% of the randomly timed glacial intervals correlate better with the LIP area in the tropics. With an associated p-value of 0.28, the null hypothesis that glacial intervals do not correlate to LIP area in the tropics cannot be rejected. Taking this approach, none of the positive or negative correlations that emerge between the LIP area scenarios are statistically significant.

@@

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic Snowball Earth events, since ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until sufficient pCO$_2$ builds up in the absence of silicate weathering to drive deglaciation. Nevertheless, the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? Using the paleomagnetic pole of \citet{Denyszyn2009a}, we reconstruct the paleolatitude of the Franklin LIP at the time of emplacement, and find that $\sim$99.7\% (or $\sim$2.6~Mkm$^{2}$) of the LIP erupted within 15\textdegree\xspace of the equator. This Franklin LIP tropical area at the time of emplacement is approximately equivalent to the Cenozoic peak, and is smaller than the other Phanerozoic peaks (Fig. \ref{fig:LIP_area}C). The ca. 1109~Ma Umkondo is another Precambrian LIP that is constrained to have erupted in the tropics, although it is not known to be associated with any glaciation (no glacial deposits are found within the contemporaneous Midcontinent Rift basin; \citealp{Swanson-Hysell2018a}). We reconstruct the paleolatitude of the Umkondo LIP at the time of emplacement using the paleomagnetic pole of \citet{Swanson-Hysell2015b}, and find that effectively all of the LIP (or $\sim$2.0~Mkm$^{2}$) erupted within the tropics, an area that is slightly smaller than that for the Franklin LIP (Fig. \ref{fig:LIP_area}C). Together, these results indicate that the tropical area of the Franklin LIP, when compared to that of Phanerozoic as well as other Precambrian LIPs, did not have a singularly large area in the tropics. Given that similar peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes must have been at play in drawing down atmospheric CO$_2$ in the lead-up to the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with arc accretion within terranes of the present-day Arabian-Nubian shield. An increase in global weatherability associated with the emplacement of the Franklin LIP in the tropics could have been a contributing factor in addition to other such processes. The temporal overlap between Franklin LIP eruptions \citep{Denyszyn2009a} and the initiation of Sturtian glaciation \citep{Macdonald2010a, MacLennan2018a} remains compelling. This overlap continues to suggest that other aspects of LIP emplacement, such as the injection of sulfur aerosols in the stratosphere \citep{Macdonald2017a}, could have played a role in the initiation of low-latitude glaciation. The latitude of such aerosol injections is important for their temporary effect on albedo which is maximized at low latitude, and their presence at high concentrations is pre-conditioned on eruption through sedimentary basins hosting evaporite deposits \citep{Macdonald2017a}. Overall, the conclusion that tropical LIP area associated with the Franklin is not uniquely high and therefore not a sole driver of Snowball Earth onset is consistent with the results of the Phanerozoic analysis. LIP area and its associated impact on global weatherability is not the fundamental control on Earth's long-term climate state.

@@

\begin{table}[h!]
\scriptsize
\caption{Phanerozoic large igneous provinces (and the Franklin).}
\resizebox{\textwidth}{!}{
\begin{tabular}{lc>{\raggedright}p{4cm}cc>{\raggedright}p{4cm}ccl}
\toprule
\textbf{name} & \textbf{age} & \textbf{age ref.} & \textbf{original} & \textbf{present} & \textbf{present area ref.} & \textbf{present/} & \textbf{half-life}$^{1}$ & \textbf{buried?} \\
 & \textbf{(Ma)} & & \textbf{area} & \textbf{area} & & \textbf{original} & \textbf{(Myr)} & \\
 & & & \textbf{(Mkm$^{2}$)} & \textbf{(Mkm$^{2}$)} & & & & \\
\midrule
Columbia River & 16 & \cite{Kasbohm2018a} & 0.68 & 0.38 & \cite{Buchan2004a} & 0.56 & 19.2 & no \\
Afar & 30 & \cite{Courtillot2003a} & 2.05 & 0.63 & \cite{Coffin2006a} & 0.31 & 17.7 & partial \\
NAIP & 62 & \cite{Larsen2015a} & 1.07 & 0.29 & \cite{Buchan2004a, Coffin2006a} & 0.27 & 33.0 & partial \\
Deccan & 66 & \cite{Schoene2014a} & 0.83 & 0.56 & \cite{Coffin2006a} & 0.68 & 116.6 & no \\
Seychelles & 66 & \cite{Schoene2014a} & 0.46 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
Madagascar & 90 & \cite{Cucciniello2010a} & 0.63 & 0.03 & \cite{Coffin2006a} & 0.05 & 20.8 & no \\
Caribbean-Colombian & 94 & \cite{Loewen2013a} & 0.71 & 0.13 & \cite{Coffin2006a} & 0.18 & 37.6 & no \\
HALIP & 95 & \cite{Kingsbury2018a} & 3.60 & 0.15 & \cite{Moosdorf2015a} & 0.04 & 20.7 & no \\
EQUAMP & 131 & \cite{Hollanda2016a} & 0.66 & 0.01 & \cite{Hollanda2016a} & 0.01 & 20.5 & no \\
Comei & 132 & \cite{Zhu2009a} & 0.11 & - & - & - & - & no \\
Bunbury & 132 & \cite{Zhu2009a} & 0.03 & 0.00 & \cite{Thorne2014a} & 0.05 & 30.9 & no \\
Parana-Etendeka & 135 & \cite{Florisbal2014a, Almeida2018a} & 3.12 & 0.40 & \cite{Coffin2006a} & 0.13 & 45.7 & partial \\
Trap & 140 & \cite{Ernst2001a} & 0.03 & 0.00 & \cite{Ernst2001a} & 0.00 & 0.0 & no \\
NW Australia Margin & 160 & \cite{Pirajno2012a} & 0.62 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
Karoo & 183 & \cite{Burgess2015a} & 3.21 & 0.15 & de Kock compilation$^{2}$ & 0.05 & 41.3 & no \\
Ferrar & 183 & \cite{Burgess2015a} & 0.18 & - & - & - & - & no \\
CAMP & 201 & \cite{Blackburn2013a} & 11.46 & 0.23 & Marzoli and Parisio compilation$^{3}$ & 0.02 & 35.7 & partial \\
Siberia & 252 & \cite{Burgess2015b} & 3.46 & 0.47 & \cite{Coffin2006a} & 0.14 & 87.5 & no \\
Emeishan & 259 & \cite{Zhou2002a} & 0.71 & 0.06 & \cite{Coffin2006a} & 0.09 & 72.9 & no \\
Panjal-Qiangtang & 283 & \cite{Zhai2013a} & 0.11 & - & - & - & - & no \\
Tarim & 290 & \cite{Xu2014a} & 0.35 & - & - & - & - & no \\
Magdalen & 360 & \cite{Brendan_Murphy1999a} & 0.42 & - & - & - & - & no \\
Vilyui & 374 & \cite{Ricci2013a} & 1.14 & - & - & - & - & no \\
Kola-Dnieper & 380 & \cite{Arzamastsev2014a} & 5.90 & - & - & - & - & no \\
Suordakh & 450 & \cite{Khudoley2013a} & 0.02 & - & - & - & - & no \\
Kalkarindji & 511 & \cite{Jourdan2014a} & 3.54 & 0.17 & \cite{Thorne2014a} & 0.05 & 116.3 & no \\
Franklin & 720 & \cite{Denyszyn2009a} & 2.62 & 0.04 & \cite{Buchan2004a} & 0.02 & 121.8 & no \\
\bottomrule
\end{tabular}
}
\begin{tablenotes}[para,flushleft]
$^{1}$assuming exponential decay with form $N(t) = 2^{t/t_{1/2}}$.

@@

\begin{figure}[h!]
\begin{center}
	\includegraphics[width=0.9\textwidth]{Manuscript/Figures/LIP_Areas.pdf}
	\caption{\textbf{A)} Paleolatitude of emplacement (centroid) and the initial area estimate of each LIP in the compilation. Car.-Col. = Caribbean-Colombian. \textbf{B)} Total LIP area through time for the different post-emplacement models. Only the `no decay' model excludes pre-Phanerozoic LIPs. \textbf{C)} Tropical LIP area through time for the different post-emplacement models. The arrows to the right indicate reconstructed tropical LIP area at the time of emplacement for the ca. 720~Ma Franklin LIP and the ca. 1109~Ma Umkondo LIP. \textbf{D)} Heat map for the preferred post-emplacement model showing the latitudinal distribution of LIP area. \textbf{E)} Latitudinal extent of land ice away from the poles through time.}
	\label{fig:LIP_area}
\end{center}
\end{figure}

>>>>

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering. A systematic analysis of LIP area and latitude through the Phanerozoic reveals that there is no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

This chapter seeks to address two interconnected questions: 1) how unique are the peaks in low-latitude LIP area that have been proposed to be associated with climatic cooling?; and 2) how strong is the overall relationship between tropical LIP area and glaciation?

@@

Outlines of the original surface extent of LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were slightly modified from the compilation of \citet{Ernst2017a}, and emplacement ages were taken from the literature (Table \ref{tab:LIPs}). The compilation of \citet{Ernst2017a} seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the construction of the original extent polygons by \citet{Ernst2017a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude from our estimation of a representative LIP area half-life the 6 of 19 LIPs which have been partially or completely buried. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the root mean square error of the fit is low (0.14), there is variability in the half-lives for LIPs if they are fit individually with an exponential decay function, such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36~Myr' uses the best-fit half-life of 36~Myr, while `t$_{1/2}$ = 120~Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly impact the calculated LIP areas throughout the Phanerozoic, which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial by sediments of LIP volcanics that are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts as well as successful rifting in the region of the Red Sea that has resulted in burial (Fig. \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios.

@@

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and Paleozoic Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. The total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree~S to 15\textdegree~N) were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_area}B and C). We use 15\textdegree~S to 15\textdegree~N as the definition of the tropical rain belt, as those latitudes approximately correspond with precipitation greater than 1.0~m/yr in modern climatalogical data \citep{Kalnay1996a}. The consistent paleolatitudes of evaporite deposits in the subtropics demonstrate the stability of the large-scale atmospheric circulation that gives rise to intense precipitation in the tropical rain belt \citep{Evans2006a}.

@@

To assess the statistical significance of the correlation (or lack thereof) implied by the Pearson correlation coefficients, we simulated the four glacial episodes (Fig. \ref{fig:LIP_area}E) occurring at random times through the past 520~Myr, and recomputed the correlation coefficient and percent overlap between the LIP area in the tropics and the randomly timed glacial intervals for each of these 100,000 simulations (Fig. \ref{fig:LIP_correlation}). This approach accounts for the reality that spurious correlation can arise between auto-correlated data sets such as these, where each value is not independent, but is instead dependent on the previous state of the system. For the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario, 28\% of the randomly timed glacial intervals correlate better with the LIP area in the tropics. With an associated p-value of 0.28, the null hypothesis that glacial intervals do not correlate to LIP area in the tropics cannot be rejected. Taking this approach, none of the positive or negative correlations that emerge between the LIP area scenarios are statistically significant.

@@

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic Snowball Earth events, since ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until sufficient CO$_2$ builds up in the absence of silicate weathering to drive deglaciation. Nevertheless, the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? Using the paleomagnetic pole of \citet{Denyszyn2009a}, we reconstruct the paleolatitude of the Franklin LIP at the time of emplacement, and find that $\sim$99.7\% (or $\sim$2.6~Mkm$^{2}$) of the LIP erupted within 15\textdegree\xspace of the equator. This Franklin LIP tropical area at the time of emplacement is approximately equivalent to the Cenozoic peak, and is smaller than the other Phanerozoic peaks (Fig. \ref{fig:LIP_area}C). The ca. 1109~Ma Umkondo is another Precambrian LIP that is constrained to have erupted in the tropics, although it is not known to be associated with any glaciation (no glacial deposits are found within the contemporaneous Midcontinent Rift basin; \citealp{Swanson-Hysell2018a}). We reconstruct the paleolatitude of the Umkondo LIP at the time of emplacement using the paleomagnetic pole of \citet{Swanson-Hysell2015b}, and find that effectively all of the LIP (or $\sim$2.0~Mkm$^{2}$) erupted within the tropics, an area that is slightly smaller than that for the Franklin LIP (Fig. \ref{fig:LIP_area}C). Together, these results indicate that the tropical area of the Franklin LIP, when compared to that of Phanerozoic as well as other Precambrian LIPs, did not have a singularly large area in the tropics. Given that similar peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes must have been at play in drawing down atmospheric CO$_2$ in the lead-up to the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with arc accretion within terranes of the present-day Arabian-Nubian Shield \citep{Park2018a}. An increase in global weatherability associated with the emplacement of the Franklin LIP in the tropics could have been a contributing factor in addition to other such processes.

The temporal overlap between Franklin LIP eruptions \citep{Denyszyn2009a} and the initiation of Sturtian glaciation \citep{Macdonald2010a, MacLennan2018a} remains compelling. This overlap continues to suggest that other aspects of LIP emplacement, such as the injection of sulfur aerosols in the stratosphere \citep{Macdonald2017a}, could have played a role in the initiation of low-latitude glaciation. The latitude of such aerosol injections is important for their temporary effect on albedo which is maximized at low latitude, and their presence at high concentrations is pre-conditioned on eruption through sedimentary basins hosting evaporite deposits \citep{Macdonald2017a}.

Overall, the conclusion that tropical LIP area associated with the Franklin is not uniquely high and therefore not a sole driver of Snowball Earth onset is consistent with the results of the Phanerozoic analysis. LIP area and its associated impact on global weatherability is not the fundamental control on Earth's long-term climate state.

@@

\begin{table}[h!]
\scriptsize
\caption{Phanerozoic large igneous provinces (and the Franklin).}
\resizebox{\textwidth}{!}{
\begin{tabular}{lc>{\raggedright}p{4cm}cc>{\raggedright}p{4cm}ccl}
\toprule
\textbf{name} & \textbf{age} & \textbf{age ref.} & \textbf{original} & \textbf{present} & \textbf{present area ref.} & \textbf{present/} & \textbf{half-life}$^{1}$ & \textbf{buried?} \\
 & \textbf{(Ma)} & & \textbf{area} & \textbf{area} & & \textbf{original} & \textbf{(Myr)} & \\
 & & & \textbf{(Mkm$^{2}$)} & \textbf{(Mkm$^{2}$)} & & & & \\
\midrule
Columbia River & 16 & \cite{Kasbohm2018a} & 0.68 & 0.38 & \cite{Buchan2004a} & 0.56 & 19.2 & no \\
 & & & & & & & & \\
Afar & 30 & \cite{Courtillot2003a} & 2.05 & 0.63 & \cite{Coffin2006a} & 0.31 & 17.7 & partial \\
 & & & & & & & & \\
NAIP & 62 & \cite{Larsen2015a} & 1.07 & 0.29 & \cite{Buchan2004a, Coffin2006a} & 0.27 & 33.0 & partial \\
 & & & & & & & & \\
Deccan & 66 & \cite{Schoene2014a} & 0.83 & 0.56 & \cite{Coffin2006a} & 0.68 & 116.6 & no \\
 & & & & & & & & \\
Seychelles & 66 & \cite{Schoene2014a} & 0.46 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
 & & & & & & & & \\
Madagascar & 90 & \cite{Cucciniello2010a} & 0.63 & 0.03 & \cite{Coffin2006a} & 0.05 & 20.8 & no \\
 & & & & & & & & \\
Caribbean-Colombian & 94 & \cite{Loewen2013a} & 0.71 & 0.13 & \cite{Coffin2006a} & 0.18 & 37.6 & no \\
 & & & & & & & & \\
HALIP & 95 & \cite{Kingsbury2018a} & 3.60 & 0.15 & \cite{Moosdorf2015a} & 0.04 & 20.7 & no \\
 & & & & & & & & \\
EQUAMP & 131 & \cite{Hollanda2016a} & 0.66 & 0.01 & \cite{Hollanda2016a} & 0.01 & 20.5 & no \\
 & & & & & & & & \\
Comei & 132 & \cite{Zhu2009a} & 0.11 & - & - & - & - & no \\
 & & & & & & & & \\
Bunbury & 132 & \cite{Zhu2009a} & 0.03 & 0.00 & \cite{Thorne2014a} & 0.05 & 30.9 & no \\
 & & & & & & & & \\
Parana-Etendeka & 135 & \cite{Florisbal2014a, Almeida2018a} & 3.12 & 0.40 & \cite{Coffin2006a} & 0.13 & 45.7 & partial \\
 & & & & & & & & \\
Trap & 140 & \cite{Ernst2001a} & 0.03 & 0.00 & \cite{Ernst2001a} & 0.00 & 0.0 & no \\
 & & & & & & & & \\
NW Australia Margin & 160 & \cite{Pirajno2012a} & 0.62 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
 & & & & & & & & \\
Karoo & 183 & \cite{Burgess2015a} & 3.21 & 0.15 & de Kock compilation$^{2}$ & 0.05 & 41.3 & no \\
 & & & & & & & & \\
Ferrar & 183 & \cite{Burgess2015a} & 0.18 & - & - & - & - & no \\
 & & & & & & & & \\
CAMP & 201 & \cite{Blackburn2013a} & 11.46 & 0.23 & Marzoli and Parisio compilation$^{3}$ & 0.02 & 35.7 & partial \\
 & & & & & & & & \\
Siberia & 252 & \cite{Burgess2015b} & 3.46 & 0.47 & \cite{Coffin2006a} & 0.14 & 87.5 & no \\
 & & & & & & & & \\
Emeishan & 259 & \cite{Zhou2002a} & 0.71 & 0.06 & \cite{Coffin2006a} & 0.09 & 72.9 & no \\
 & & & & & & & & \\
Panjal-Qiangtang & 283 & \cite{Zhai2013a} & 0.11 & - & - & - & - & no \\
 & & & & & & & & \\
Tarim & 290 & \cite{Xu2014a} & 0.35 & - & - & - & - & no \\
 & & & & & & & & \\
Magdalen & 360 & \cite{Brendan_Murphy1999a} & 0.42 & - & - & - & - & no \\
 & & & & & & & & \\
Vilyui & 374 & \cite{Ricci2013a} & 1.14 & - & - & - & - & no \\
 & & & & & & & & \\
Kola-Dnieper & 380 & \cite{Arzamastsev2014a} & 5.90 & - & - & - & - & no \\
 & & & & & & & & \\
Suordakh & 450 & \cite{Khudoley2013a} & 0.02 & - & - & - & - & no \\
 & & & & & & & & \\
Kalkarindji & 511 & \cite{Jourdan2014a} & 3.54 & 0.17 & \cite{Thorne2014a} & 0.05 & 116.3 & no \\
 & & & & & & & & \\
Franklin & 720 & \cite{Denyszyn2009a} & 2.62 & 0.04 & \cite{Buchan2004a} & 0.02 & 121.8 & no \\
\bottomrule
\end{tabular}
}
\begin{tablenotes}[para,flushleft]
$^{1}$assuming exponential decay with form $N(t) = 2^{t/t_{1/2}}$.

@@

\begin{figure}[h!]
\begin{center}
	\includegraphics[width=0.8\textwidth]{Manuscript/Figures/LIP_Areas.pdf}
	\caption{\textbf{A)} LIPs included in this analysis. The size of each circle reflects the initial area estimate of each LIP. The + indicates the timing and absolute paleolatitude of the centroid of each LIP at the time of emplacement. Car.-Col. = Caribbean-Colombian. \textbf{B)} Total LIP area through time for the different post-emplacement models. Only the `no decay' model excludes pre-Phanerozoic LIPs. \textbf{C)} Tropical LIP area through time for the different post-emplacement models. The arrows to the right indicate reconstructed tropical LIP area at the time of emplacement for the ca. 720~Ma Franklin LIP and the ca. 1109~Ma Umkondo LIP. \textbf{D)} Heat map for the preferred post-emplacement model showing the latitudinal distribution of LIP area. \textbf{E)} Latitudinal extent of land ice away from the poles through time \citep{Macdonald2018a}.}
	\label{fig:LIP_area}
\end{center}
\end{figure}
